---
title: "193DS_file"
author: Mariko Bronson
date: 06-08-2025
format:
  html:
    toc: true
    execute:
      message: false 
      warning: false 
---

https://github.com/MarikoBron/ENVS-193DS_spring-2025_final.git

```{r data-and-packages}
#installing packages
library(tidyverse)
library(here)
library(janitor)
library(flextable)
library(readxl)
library(ggplot2)
library(colorspace)
library(dplyr) 
library(lterdatasampler)
library(DHARMa)
library(MuMIn)
library(ggeffects)

#create object sst 
sst <- read_csv(
  here("data", "SST.csv"))

#create object nest_boxes
nest_boxes <- read_csv(
  here("data","nest_boxes.csv"))

#create object screentime from my data
screentime <- read_xlsx(
  here("data", "193DS_screentime_updated.xlsx"),
  sheet = "Sheet1")

```

# Problem 1: Research Writing

## a. Transparent statistical methods

In part 1, they used a Pearson's correlation to find how the variables are related to each other. In part 2, they used an ANOVA to find whether the means of multiple groups.

## b. More information needed

In part 2, along with the ANOVA, he should include the effect size and the Turkey HSD. The ANOVA tells you whether there is a difference in the means, and the effect size tells you how large of an effect sources have on the difference. This is important because it helps you find out what might be causing these differences. The ANOVA tells you whether differences are present, and the Turkey HSD will tell you which groups are different and by how much.

## c. Suggestions for rewriting

Our research suggests that there is a positive correlation between annual total nitrogen load (kg/year) and distance from headwater (km) in the San Joaquin Delta River(Pearson's correlation, r = correlation coefficient, p = 0.03, $\\alpha$ = 0.05) with a large($\\eta^2$ = effect size) effect of distance. Sources(urban land, atmospheric depostion, fertilizer, wastewater treatment, and grasslands) differed in average nitrogen load (one-way ANOVA, F(4, df) = test statistic, p = 0.02, $\\alpha$ = 0.05). We found that fertilizer had a higher nitrogen load than atmospheric deposition (Tukey HSD, p = significance, 95% confidence interval: \[lower CI, upper CI\] kg/year), and atmospheric deposition had a higher nitrogen load than wastewater treatment (Tukey HSD, p = significance, 95% confidence interval: \[lower CI, upper CI\] kg/year).

# Problem 2:

## a. Cleaning and summarizing

```{r cleaning-and-summarizing}

#create sst_clean
sst_clean <- sst |> 
  select(date, temp) |> 
#select 2018-2023
  filter(year(date) >= 2018 & year(date) <= 2023) |> 
# create year and month columns as factors
  mutate(
    year = factor(year(date)),
    month = factor(month(date, label = TRUE))) |> 
#delete date column
  select(-date) |> 
#calculate mean monthly sst
  group_by(year, month) |> 
  summarise(
    mean_monthly_sst = mean(temp)) |> 
#ungroup 
  ungroup()

#display 5 random points
slice_sample(sst_clean, n = 5)
#check structure
str(sst_clean)
```

## b. Visualize the data

```{r recreating-graph}

#start ggplot, choose axes and group by color
ggplot(sst_clean,
       aes(x = month,
           y = mean_monthly_sst,
           color = year,
           group = year)) +
#adding points
   geom_point() +
#adding lines
  geom_line() +
#adding labels
  labs(x = "Month",
       y = "Mean Monthly Sea Surface Temperature(°C)",
       color = "Year") +
#cleaner background, scootch the legend
   theme_minimal() +
    theme(
    panel.grid = element_blank(),
    panel.border = element_rect(fill = NA),
    legend.position = c(0.1,0.75)) +
  
#cute colors
scale_color_discrete_sequential(palette = "Purp") 

```

# Problem 3: Data Analysis

## a. Response variable

Response variable columns are labeled with abbreviated conditions of occupancy in the nest boxes(swift parrot, common starling, empty, and tree martin) where 1s refer to the condition being found to be true and 0s refer to the condition being found to be false. For example, a 1 under sp means a swift parrot was there, and a 0 under sp means a swift parrot was not there.

## b. Purpose of study

The main difference is that the nest boxes are intended to provide habitat for swift parrots, and the study means to investigate whether nest boxes are specialized enough to be a worthwhile and successful restoration tool. When birds besides swift parrots are found in the boxes, it indicates that the nest boxes are not working as intended.

## c. Difference in "seasons"

The seasons refer to the years 2016 and 2019, which are important because mass tree flowering events during these years triggered breeding events at the study site. The nest boxes were installed in 2016 to test the strategy of deploying boxes where breeding is most probable, and were checked in 2019 after to test permanent nest boxes that have been there for several years.

## d. Table of models

| Model number | Season | Distance to Forest Edge | Model Description |
|:---------------:|:---------------:|:---------------:|:--------------------|
| 0 |  |  | no predictors (null model) |
| 1 | X | X | all predictors (saturated model) |
| 2 |  | X | distance to forest edge |
| 3 | X |  | season |

## e. Run the models

```{r cleandat, echo = FALSE}
nest_boxes_clean <- clean_names(nest_boxes)
```

```{r models}

# model 0: null model
model0 <- glm(
  sp ~ 1, # formula
  data = nest_boxes_clean, # data frame
  family = "binomial" #distribution
)

# model 1: saturated model
model1 <- glm(
  sp ~ season + edge_distance,
  data = nest_boxes_clean, 
  family = "binomial"
)

# model 2: distance to forest edge
model2 <- glm(
  sp ~ edge_distance,
  data = nest_boxes_clean,
  family = "binomial"
)

# model 3: season
model3 <- glm(
  sp ~ season,
  data = nest_boxes_clean,
  family = "binomial"
)

```

## f. Check the diagnostics

```{r diagnostics}

#create 4 plots of residual diagnostics
 plot(
  simulateResiduals(model0)) 
  
 plot(
  simulateResiduals(model1))  
  
 plot(
  simulateResiduals(model2)) 
  
 plot(
  simulateResiduals(model3))

```

## g. Select the best model

```{r model-selection}

#create AIC model selection table
AICc(model0,
          model1, 
          model2, 
          model3)

```

The best model that predicts the presence of a swift parrot as determined by Akaike’s Information Criterion (AIC) includes the season and distance from the edge.

## h. Visualize the model predictions

```{r visualize}

#create object for predictions
mod_preds <- ggpredict(model1,
                       terms = c("edge_distance [all]", "season"))
#plot model predictions
ggplot(mod_preds, #create plot frame
       aes(
         x = x,
       y = predicted,
       color = group)) + #choosing axes and coloring by group
  labs(
    x = "Distance from the forest edge (m)",
    y = "Probability of box occupancy", #labeling axes
    color = "season",
    fill = "season"  # fix legend to match underlying data points 
  ) +
    geom_point(size = 1, #add transparent underlying data points
               alpha = 0.4) +
  geom_ribbon(data = mod_preds, #add ribbon with predictions and CI
              aes(
                x = x,
                y = predicted,
                ymin = conf.low,
                ymax = conf.high, 
      fill = group  #coloring ribbons by group
              ),
              alpha = 0.4) +
  geom_line(data = mod_preds, #add line with predictions
            aes(
              x = x,
              y = predicted 
            ))  +
  theme_minimal() + #clean up chart
  theme(
    panel.grid = element_blank(), #clean up chart
    panel.border = element_rect(fill = NA)
  ) +
  
  scale_color_discrete_diverging(palette = "Tropic") + 
  scale_fill_discrete_diverging(palette = "Tropic") #choose colors

```

## i. Write a caption for your figure

### Figure 1: Probability of box occupancy decreases as distance from forest edge increases. 

Predictions and CI from the best models of occupancy of nest boxes by Swift Parrots, which are the target species of restoration efforts. Data was collected during mass breeding events in 2016, when nest boxes were first installed, and in 2019 after 3 years in place. Data sourced from: Stojanovic, Dejan et al. (2021). Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird. Dryad. <https://doi.org/10.5061/dryad.83bk3j9sb>

## j. Calculate model predictions

```{r calculate-predictions}

#finding probability of occupancy from 0-900m by season
mod_preds_95 <- ggpredict(model1, 
                       terms = c("edge_distance [0,900]", "season"))
print(mod_preds_95) #displaying results

```

## k. Interpret your results

In 2016, the predicted probability of nest box occupancy by Swift Parrots is 48% at 0 m from the forest edge and 12% at 900 m from the forest edge based on ggpredict results. In 2019, predicted probability of occupancy is 30% at 0 m from the forest edge and 6% at 900 m from the forest edge. There is a negative relationship between distance from the forest edge and nest box occupancy by Swift Parrots in both years. This could be because the nest boxes further from the forest edge are strongly preferred by Tree Martins, which is interesting because Tree Martins are subordinate competitors to both Swift Parrots and Common Starlings.

# Problem 4: **Affective and exploratory visualizations**

## a. Comparing visualizations

The visualization I made in Homework 2 does not just suffer from a severe lack of data, but is also much less pleasant and interesting to look at. I had not yet gathered enough data to find relationships between variables, so I was not using the best predictive models. The charts are bland and have a poor ink:data ratio.

There are no similarities between all three visualizations except for the fact that they all use screentime in hours as a dependent variable. Both my final visualization and my second chart focus on task to screentime relationships, though one uses tasks done and one uses tasks due.

Tasks due has a stronger relationship than tasks done on screentime, but the not by much, so I was on the right track with my second visualization. Even though I only had a couple datapoints, a negative relationship between workload and screentime was already emerging. It makes sense that I'm on my phone less when I have more things to do, which does reject my initial thought that sparked this analysis.

I did hadn't started my visual yet in week 9, but I did give myself mental feedback based on the projects I saw. While both were very well done, they were very similar to examples we were given, and I personally found one to be a little overwhelming to look at and the other a bit underwhelming. I decided I wanted to focus on making a visualization that is original, easy to understand, and a little funky.
