---
title: "193DS_file"
author: Mariko Bronson
date: 06-08-2025
format:
  html:
    toc: true
    execute:
      message: false 
      warning: false 
---

https://github.com/MarikoBron/ENVS-193DS_spring-2025_final.git

```{r data-and-packages}
#installing packages
library(tidyverse)
library(here)
library(janitor)
library(flextable)
library(readxl)
library(ggplot2)
library(colorspace)
library(dplyr) 
library(lterdatasampler)
library(DHARMa)

#create object sst 
sst <- read_csv(
  here("data", "SST.csv"))

#create object nest_boxes
nest_boxes <- read_csv(
  here("data","nest_boxes.csv"))

#create object screentime from my data
screentime <- read_xlsx(
  here("data", "193DS_screentime_updated.xlsx"),
  sheet = "Sheet1")

```

# Problem 1: Research Writing

## a. Transparent statistical methods

In part 1, they used a Pearson's correlation to find how the variables are related to each other. In part 2, they used an ANOVA to find whether the means of multiple groups.

## b. More information needed

In part 2, along with the ANOVA, he should include the effect size and the Turkey HSD. The ANOVA tells you whether there is a difference in the means, and the effect size tells you how large of an effect sources have on the difference. This is important because it helps you find out what might be causing these differences. The ANOVA tells you whether differences are present, and the Turkey HSD will tell you which groups are different and by how much.

## c. Suggestions for rewriting

Our research suggests that there is a positive correlation between annual total nitrogen load (kg/year) and distance from headwater (km) in the San Joaquin Delta River(Pearson's correlation, r = correlation coefficient, p = 0.03, $\\alpha$ = 0.05) with a large($\\eta^2$ = effect size) effect of distance. Sources(urban land, atmospheric depostion, fertilizer, wastewater treatment, and grasslands) differed in average nitrogen load (one-way ANOVA, F(4, df) = test statistic, p = 0.02, $\\alpha$ = 0.05). We found that fertilizer had a higher nitrogen load than atmospheric deposition (Tukey HSD, p = significance, 95% confidence interval: \[lower CI, upper CI\] kg/year), and atmospheric deposition had a higher nitrogen load than wastewater treatment (Tukey HSD, p = significance, 95% confidence interval: \[lower CI, upper CI\] kg/year).

# Problem 2:

## a. Cleaning and summarizing

```{r cleaning-and-summarizing}

#create sst_clean
sst_clean <- sst |> 
  select(date, temp) |> 
#select 2018-2023
  filter(year(date) >= 2018 & year(date) <= 2023) |> 
# create year and month columns as factors
  mutate(
    year = factor(year(date)),
    month = factor(month(date, label = TRUE))) |> 
#delete date column
  select(-date) |> 
#calculate mean monthly sst
  group_by(year, month) |> 
  summarise(
    mean_monthly_sst = mean(temp)) |> 
#ungroup 
  ungroup()

#display 5 random points
slice_sample(sst_clean, n = 5)
#check structure
str(sst_clean)
```

## b. Visualize the data

```{r recreating-graph}

#start ggplot, choose axes and group by color
ggplot(sst_clean,
       aes(x = month,
           y = mean_monthly_sst,
           color = year,
           group = year)) +
#adding points
   geom_point() +
#adding lines
  geom_line() +
#adding labels
  labs(x = "Month",
       y = "Mean Monthly Sea Surface Temperature(°C)",
       color = "Year") +
#cleaner background, scootch the legend
   theme_minimal() +
    theme(
    panel.grid = element_blank(),
    panel.border = element_rect(fill = NA),
    legend.position = c(0.1,0.75)) +
  
#cute colors
scale_color_discrete_sequential(palette = "Purp") 

```

# Problem 3: Data Analysis

## a. Response variable

Response variable columns are labeled with abbreviated conditions of occupancy in the nest boxes(swift parrot, common starling, empty, and tree martin) where 1s refer to the condition being found to be true and 0s refer to the condition being found to be false. For example, a 1 under sp means a swift parrot was there, and a 0 under sp means a swift parrot was not there.

## b. Purpose of study

The main difference is that the nest boxes are intended to provide habitat for swift parrots, and the study means to investigate whether nest boxes are specialized enough to be a worthwhile and successful restoration tool. When birds besides swift parrots are found in the boxes, it indicates that the nest boxes are not working as intended.

## c. Difference in "seasons"

The seasons refer to the years 2016 and 2019, which are important because mass tree flowering events during these years triggered breeding events at the study site. The nest boxes were installed in 2016 to test the strategy of deploying boxes where breeding is most probable, and were checked in 2019 after to test permanent nest boxes that have been there for several years.

## d. Table of models

| Model number | Season | Distance to Forest Edge | Model Description |
|:--------------:|:--------------:|:---------------:|:----------------------|
| 0 |  |  | no predictors (null model) |
| 1 | X | X | all predictors (saturated model) |
| 2 |  | X | distance to forest edge |
| 3 | X |  | season |

## e. Run the models

```{r models}

# model 0: null model
model0 <- glm(
  sp ~ 1, # formula
  data = nest_boxes, # data frame
  family = "binomial" #distribution
)

# model 1: saturated model
model1 <- glm(
  sp ~ season + `edge distance`,
  data = nest_boxes, 
  family = "binomial"
)

# model 2: distance to forest edge
model2 <- glm(
  sp ~ `edge distance`,
  data = nest_boxes,
  family = "binomial"
)

# model 3: season
model3 <- glm(
  sp ~ season,
  data = nest_boxes,
  family = "binomial"
)

```

## f. Check the diagnostics

```{r diagnostics}

plot(
  simulateResiduals(nest_boxes)
)

```
